Bridging the foundational gap: updating algebraic geometry in face of current challenges regarding formalizability, constructivity and predicativity

### 2021-11-20 slides-dagstuhl2021.tex
Slides for Dagstuhl 2021 (WIP)

\usepackage{pgfpages}
\setbeameroption{show notes on second screen}
\setbeamertemplate{note page}[plain]
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}

@@

\usepackage{graphbox}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage[normalem]{ulem}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning}
\hypersetup{colorlinks=true}

@@

\title{A general Nullstellensatz for generalised spaces}
\author{Ingo Blechschmidt}
\date{April 11th, 2019}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

\newcommand{\bignumber}[1]{
  \renewcommand{\insertenumlabel}{#1}\scalebox{1.5}{\usebeamertemplate{enumerate item}}
}
\newcommand{\bigheart}{\includegraphics{heart}}

@@

\setbeamertemplate{headline}{%
  \begin{beamercolorbox}[wd=\paperwidth,ht=2.25ex]{}%
    \insertsectionnavigationhorizontal{\paperwidth}{}{}%
  \end{beamercolorbox}%
  \vskip0pt%
}

\setbeamertemplate{frametitle}{%
  \vskip0.4em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
  %   \usebeamercolor[fg]{item}{\textbf{{\Large \insertframetitle}}}
    \begin{tikzpicture}
      \def\R{8pt}
      \node (title) {\hil{\large\insertframetitle}};
      \begin{pgfonlayer}{background}
        \draw[decorate, very thick, draw=mypurple]
          ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
          ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
          ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
          ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
          cycle;
      \end{pgfonlayer}
    \end{tikzpicture}
  \end{beamercolorbox}%
  \vskip-0.6em%
}

@@

@@

  \bigskip
  \[ \xymatrix{
    & \text{$\ul{\TT}/U_\TT$ proves $\sigma$} \\
    \text{$\sigma$ holds for $U_\TT$} \ar@/^1.1pc/@{<=>}[ur] &&
    \text{$\sigma$ holds for $M$} \ar@/_1.1pc/@{<=}[ul]
  } \]
  \bigskip

  \begin{tikzpicture}
    \def\R{8pt}
    \node (title) {\hil{A general Nullstellensatz for generalised spaces}};
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

@@

  Category Theory 2019 in Edinburgh \\
  July 9th, 2019
  \bigskip

  Ingo Blechschmidt \\
  Università di Verona
  \par

  \jnote{1}{\textbf{This talk in a nutshell:}

  There is a certain ring, the \emph{generic ring}, which is special in that
  it is \emph{conservative}: It has exactly those properties which are shared
  by any ring whatsoever. One could argue that it is this ring which we're
  implicitly referring to when we utter the phrase ``Let~$R$ be a ring.''.

  A surprising observation is that the generic ring is a field. This is not a
  contradiction because conservativity only pertains to properties which can be
  put in a certain syntactic form, dubbed \emph{geometric sequents}, and the
  field condition is not of that form.

  As a consequence, when verifying a geometric sequent for all rings, we may
  freely assume the field condition: Our proof will apply to the generic ring,
  because this ring happens to be a field, and its conclusion will spread to
  any ring whatsoever thanks to its syntactic form.

  An analogous story can be told for the generic group, the generic vector
  space and quite generally for the generic model of any \emph{geometric
  theory}. It is an interesting task to discover properties of such
  generic models. This talk presents a source of such properties, the
  \emph{topos-theoretic Nullstellensatz}, which is
  universal in the sense that any such property is a logical consequence of the
  Nullstellensatz.}
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\hspace*{-1.7cm}\includegraphics[height=\paperheight,trim=0 0 5.1cm 0,clip]{arthurs-seat-4}\end{minipage}}
\begin{frame}[plain]%
  \vspace*{\paperheight}\vspace*{-2.35cm}\hspace*{-1cm}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-7}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-0}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-5}%
  \includegraphics[width=\paperwidth/4]{arthurs-seat-9}%
  \jnote{1}{This talk owes a special debt to Alexander Oldenziel, who
  repeatedly suggested that the topic of this talk should be studied.

  The photos on the slide are of \emph{Arthur's Seat}, a beautiful extinct
  volcano in Edinburgh where parts of these slides have been created. It gives
  an excellent view of the city and is very much recommended for a hiking tour.
  }%
\end{frame}}
\addtocounter{framenumber}{-1}

@@

\subsection{The mystery of nongeometric sequents}

\begin{frame}{The mystery of nongeometric sequents}
  \jnote{1}{A \emph{geometric sequent} is a syntactic expression of the
  form~$(\varphi \seq{x_1\?X_1,\ldots,x_n\?X_n} \psi)$,
  where~$x_1\?X_1,\ldots,x_n\?X_n$ is a list of variable declarations,
  the~$X_i$ ranging over the available sorts, and~$\varphi$ and~$\psi$ are
  \emph{geometric formulas}. Such a sequent is read as ``in the
  context~$x_1\?X_1,\ldots,x_n\?X_n$, $\varphi$ entails $\psi$''.
  Geometric formulas are built from atomic propositions (equality and the
  available relation symbols) using the connectives~$\top$, $\bot$, $\wedge$, $\bigvee$
  (set-indexed disjunction) and~$\exists$. Geometric formulas may not
  contain~$\neg$, $\Rightarrow$, $\forall$.

  There is a notion of a \emph{model} of a geometric theory in
  any given topos. For instance, a ring in the usual sense is a model of the
  theory of rings in the topos~$\Set$. The structure sheaf~$\O_X$ of a scheme~$X$ is a
  model in the topos~$\Sh(X)$ of set-valued sheaves over~$X$.

  With \emph{topos} we mean Grothendieck topos, and as metatheory we use a
  constructive but impredicative flavour of English (which could be formalised
  by what is supported by the internal language of elementary toposes with an NNO).
  However the Nullstellensatz presented later makes no use of the
  subobject classifier, hence the results can likely be generalised to hold in
  a predicative metatheory or to hold for arithmetic universes.}

  \jnote{2-}{Among all models in any topos, the \emph{universal} or
  \emph{generic} one is special. It enjoys the universal property that any
  model in any topos can be obtained from it by pullback along an essentially
  unique geometric morphism. It is intriguing from a logical point of view
  because it has exactly those properties which are shared by any model in any
  topos.

  One could argue, with a certain amount of success, that the generic model of
  the theory of rings is what a mathematician implicitly refers to when she
  utters the phrase ``Let~$R$ be a ring''. This point of view is fundamental to
  the slogan \emph{continuity is geometricity}, as expounded for instance in
  \fixedhref{http://www.cs.bham.ac.uk/~sjv/GeoAspects.pdf}{Continuity and
  geometric logic} by Steve Vickers.}

  \jnote{3-}{Crucially, the conservativity statement only pertains to
  properties which can be put as geometric sequents. Generic models may have
  additional nongeometric properties. Because conservativity does not apply to
  them, they are not shared by all models in all toposes -- but any consequences
  which can be put as geometric sequents are.

  For instance, if we want to verify a geometric sequent for all local rings,
  we may freely use the displayed field axiom.
  Hence one reason why these nongeometric sequents are interesting is because
  they provide us with new reduction strategies (proofs by ``without loss of
  generality'').}

  Let~$\TT$ be a \explain{geometric theory}{1-}{red!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
    \small
    sorts,
    function symbols,
    relation symbols,
    geometric sequents as axioms
  \end{minipage}}, for instance the
  \explain{theory of rings}{1-}{blue!20}{-1cm}{0cm}{0cm}{\begin{minipage}{4.3cm}
    \small
    \begin{tabbing}
      fun. symb.: \= \kill
      sorts: \> $R$ \\
      fun. symb.: \> $0$, $1$, $-$, $+$, $\cdot$ \\
      % rel. symb.: \> \emph{none} \\
      axioms: \> $(\top \vdash_{x,y:R} x y = y x)$, \ldots
    \end{tabbing}
  \end{minipage}}.

  \vspace*{1.8cm}

  \begin{center}
    \begin{tikzpicture}[ultra thick, node distance=7mm]
      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$};
      \node[regular polygon, regular polygon sides=5, draw=orange!80,
      fill=orange!20, right=of b, rounded corners=1pt, inner sep=0cm] (c) {$\O_X$};
      \node[star, rounded corners=1pt, star points=10, inner sep=0cm, draw=purple!80, fill=purple!20, right=of c] {$U_\TT$};
    \end{tikzpicture}
  \end{center}

  \vspace*{-0.5cm}
  \pause

  \justifying
  \textbf{Theorem.} There is a \hil{generic model}~$U_\TT$. It is
  \hil{conservative} in that
  for any \explainstub{geometric sequent}{yellow!70}~$\sigma$ the following notions coincide:
  \vspace*{-1.2em}
  \begin{enumerate}
    \item The sequent~$\sigma$ holds for~$U_\TT$.
    \item The sequent~$\sigma$ holds for any~$\TT$-model in any
    \explainstub{topos}{purple!20}.
    \item The sequent~$\sigma$ is provable modulo~$\TT$.
  \end{enumerate}

  \pause

  \justifying
  %\textbf{Main observation.} The generic model can validate nongeometric
  %sequents which are not provable modulo~$\TT$ and which do not hold for
  %arbitrary models in arbitrary toposes.

  \textbf{Observation (Kock).} The generic local ring is a field:
  \vspace*{-0.2cm}
  \[ (x = 0 \Rightarrow \bot) \vdash_{x:R} (\exists y\?R\_ xy = 1) \]
\end{frame}

\subsection{Construction of the generic model}

\begin{frame}{Construction of the generic model}
  The generic model is \hil{not} the same as \ldots
  \begin{itemize}
    \item the \hil{initial model} (think~$\ZZ$) or
    \item the \hil{free model on one generator} (think~$\ZZ[X]$).
  \end{itemize}
  Set-based models are \hil{too inflexible}.

  \textbf{Definition.} The \hil{syntactic site}~$\C_\TT$ has \ldots
  \begin{enumerate}
    \item objects: \explainstub{formulas-in-contexts}{red!20} $\{ x_1\?X_1, \ldots, x_n\?X_n\_ \varphi \}$
    \item morphisms: eqv. classes of \explainstub{provably functional
    formulas}{red!30}
    \item coverings: \explainstub{provably jointly surjective
    families}{purple!30}
  \end{enumerate}
  The topos of sheaves over~$\C_\TT$ is the \hil{classifying
  topos}~$\Set[\TT]$. The generic model interprets a sort~$X$ by the sheaf~$\yon\{x\?X\_\top\}$.

  \jnote{1}{In case the theory~$\TT$ is a Horn theory (for instance if it is
  an equational theory), the \emph{term algebra} (the set of terms in the
  empty context modulo provable equality) is a model of~$\TT$. While the
  term algebra does enjoy some nice categorical properties, it is in general
  \emph{not} the generic model.

  For instance, if~$\TT$ is the theory of rings, then the initial model
  is~$\ZZ$. This model validates some geometric sequents which are not
  validated by all rings, for instance~$(x^2 = 0 \seq{x:R} x = 0)$ or $(1 = 0
  \vdash \bot)$.

  In general, the generic model cannot be realised as a set-based model (with
  a set for each sort, a map for each function symbol and so on). Sets are too
  constant for this purpose; the flexibility of sheaves (``variable sets'') is
  required: The generic model lives in the topos of set-valued
  sheaves over~$\C_\TT$, called the \emph{classifying topos} of~$\TT$.}

  \jnote{2}{The notation in the definition of the objects of~$\C_\TT$ is meant
  to be suggestive, but should not be taken literally. The objects are
  syntactic expressions consisting of a context and a geometric formula in that
  context.

  A morphism~$A = \{\vec x\_ \varphi\} \to \{\vec y\_ \psi\} = B$ in~$\C_\TT$ is the
  equivalence class (modulo provable equivalence) of a geometric
  formula~$\theta$ such that~$\TT$ proves
  \begin{enumerate}
    \item ``$\theta$ is a relation on~$A \times B$'': $(\theta \seq{\vec x,
    \vec y} \varphi \wedge \psi)$
    \item ``$\theta$ is total'': $(\varphi \seq{\vec x} \exists \vec y\_
    \theta)$
    \item ``$\theta$ is single-valued'': $(\theta \wedge \theta[\vec y'/\vec y]
    \seq{\vec x,\vec y,\vec y'} \vec y = \vec y')$
  \end{enumerate}

  \vspace*{-0.6em}
  A family~$(\{ \vec x_i\_ \varphi_i \} \xrightarrow{[\theta_i]} \{ \vec y\_
  \psi \})_i$ is a covering iff~$\TT$ proves~$(\psi \seq{\vec y} \bigvee_i
  \exists \vec x_i\_ \theta_i)$.

  The slides experiment with using the symbol~``$\yon$'' for the Yoneda
  embedding~$\C_\TT \to \Set[\TT]$, as in \fixedhref{http://www.math.jhu.edu/~eriehl/elements.pdf}{Elements
  of~$(\infty,1)$-category theory} by Emily Riehl and Dominic Verity.}

  \jnote{3}{The special case that the generic model of a theory~$\TT$ can be realised as
  a model in~$\Set$ occurs iff~$\TT$ is Morita-equivalent to the
  empty theory, that is, iff~$\TT$ has exactly one model in any topos.

  The special case that there exists at least some conservative~$\TT$-model
  in~$\Set$ occurs iff~$\TT$ has a conservative geometric expansion to a theory
  which is Morita-equivalent to the empty theory.}
\end{frame}

\subsection{Going internal}

>>>>

\usepackage{pgfpages}
%\setbeameroption{show notes on second screen}
%\setbeamertemplate{note page}[plain]
%\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\justifying\footnotesize#2\par}}}
\newcommand{\jnote}[2]{}

@@

\usepackage{graphbox}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage[normalem]{ulem}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing,positioning}
\hypersetup{colorlinks=true}

\newcommand*\circled[1]{\tikz[baseline=(char.base)]{%
  \node[shape=circle,draw,inner sep=1pt] (char) {#1};}}

\DeclareFontFamily{U}{bbm}{}
\DeclareFontShape{U}{bbm}{m}{n}
   {  <5> <6> <7> <8> <9> <10> <12> gen * bbm
      <10.95> bbm10%
      <14.4>  bbm12%
      <17.28><20.74><24.88> bbm17}{}
\DeclareFontShape{U}{bbm}{m}{sl}
   {  <5> <6> <7> bbmsl8%
      <8> <9> <10> <12> gen * bbmsl
      <10.95> bbmsl10%
      <14.4> <17.28> <20.74> <24.88> bbmsl12}{}
\DeclareFontShape{U}{bbm}{bx}{n}
   {  <5> <6> <7> <8> <9> <10> <12> gen * bbmbx
      <10.95> bbmbx10%
      <14.4> <17.28> <20.74> <24.88> bbmbx12}{}
\DeclareFontShape{U}{bbm}{bx}{sl}
   {  <5> <6> <7> <8> <9> <10> <10.95> <12> <14.4> <17.28>%
      <20.74> <24.88> bbmbxsl10}{}
\DeclareFontShape{U}{bbm}{b}{n}
   {  <5> <6> <7> <8> <9> <10> <10.95> <12> <14.4> <17.28>%
      <20.74> <24.88> bbmb10}{}
\DeclareMathAlphabet{\mathbbm}{U}{bbm}{m}{n}
\SetMathAlphabet\mathbbm{bold}{U}{bbm}{bx}{n}

@@

\title{Bridging the foundational gap: updating algebraic geometry in
face of current challenges regarding formalizability, constructivity and
predicativity}
\author{Ingo Blechschmidt}
\date{November 23th, 2021}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\BB}{\mathbb{B}}
\newcommand{\pp}{\mathbbm{p}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

@@

\setbeamertemplate{headline}{}

\setbeamertemplate{frametitle}{%
  \vskip0.5em%
  \leavevmode%
  \begin{beamercolorbox}[dp=1ex,center]{}%
    \usebeamercolor[fg]{item}{\textbf{{\Large\insertframetitle}}}
  \end{beamercolorbox}%
  \vskip-0.2em%
}

@@

\newcommand{\bignumber}[1]{%
  \renewcommand{\insertenumlabel}{#1}\scalebox{1.2}{\!\usebeamertemplate{enumerate item}\!}
}
\newcommand{\bigheart}{\includegraphics{heart}}

\newcommand{\subhead}[1]{{\centering\textcolor{gray}{\hrulefill}\quad\textnormal{#1}\quad\textcolor{gray}{\hrulefill}\par}}

@@

  \bigskip
  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
  \bigskip

  \begin{tikzpicture}
    \def\R{8pt}
    \node (pretitle) {Bridging the foundational gap:};
    \node (title) [below of=pretitle] {\vbox{\vspace*{-0.5em}updating \hil{algebraic
    geometry} in face of \hil{current challenges} regarding
    \hil{formalizability}, \hil{constructivity} and \hil{predicativity}}};
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

@@

  Dagstuhl Seminar 20202: \\
  Geometric Logic, Constructivisation, and Automated Theorem Proving \\
  \ \\
  November 23th, 2021
  \bigskip

  Ingo Blechschmidt \\
  University of Augsburg
  \par
\end{frame}}

\begin{frame}{Algebraic geometry in a nutshell}
  Turn \hil{commutative rings} into \hil{spaces}, and \hil{glue} those
  spaces.
  \bigskip

  \subhead{Examples}
  \begin{enumerate}
    \item $\Spec(k[X_1,\ldots,X_n]) = \AA^n$.
    \item $\Spec(k[X,Y]/(Y-X^2)) = \text{standard parabola}$.
    \item Gluing
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (90:1) circle (0.2);
    \end{tikzpicture}
    with
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (270:1) circle (0.2);
    \end{tikzpicture}
    along
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (90:1) circle (0.2);
      \draw [fill=white] (270:1) circle (0.2);
    \end{tikzpicture}
    yields
    $\PP^1$.
  \end{enumerate}
  \bigskip

  \subhead{Concrete results}
  Fermat's Last Theorem: For~$n \geq 3$, no positive integers satisfy
  \[ a^n + b^n = c^n. \]
\end{frame}

\begin{frame}{Transfinite methods}
  The current presentation of algebraic geometry employs:
  \begin{itemize}
    \item large structures: classes, large categories, universes, \ldots
    \item law of excluded middle
    \item axiom of choice
  \end{itemize}
  \bigskip

  \subhead{despite:}
  \begin{enumerate}
    \item subject matter (in part) very concrete
    \item computer algebra systems for computations exist
    \item constructive algebra well-developed
  \end{enumerate}
\end{frame}

@@

\subsection{The generic ring}

\begin{frame}{The generic ring}
  \visible<2->{``}Let~$R$ be a ring.\visible<2->{''} \visible<2->{-- Which ring does this phrase refer to?}

  \vspace*{-0.6em}
  \begin{center}
    \begin{tikzpicture}[ultra thick, node distance=7mm]
      \node[rectangle, rounded corners=1pt, inner sep=5pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\FF_2$};
      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of b] (c) {$\QQ[X]$};
      \node[regular polygon, regular polygon sides=6, draw=orange!80, fill=orange!40, right=of c, rounded corners=1pt] (d) {$\RR$};
      \node[regular polygon, regular polygon sides=5, draw=pink!80, fill=pink!40, right=of d, rounded corners=1pt, inner sep=0cm] (e) {$\O_X$};
      \visible<3->{\node[star, rounded corners=1pt, star points=10, inner
      sep=2pt, draw=purple!80, fill=purple!30, right=of e] {$\AA$};}
    \end{tikzpicture}
  \end{center}
  \vspace*{-0.6em}

  \pause
  \pause
  \pause

  \justifying
  \textbf{Thm.} For any$^\star$ property~$P$ of rings, the following are
  equivalent: \\[0em]
  \begin{enumerate}
    \item The \hil{generic ring}~$\AA$ has property~$P$.
    \item Every$^\star$ ring has property~$P$.
    \item The ring axioms entail property~$P$.
  \end{enumerate}
  \pause

  \textbf{Example A.} For any~$x,y,z \in \AA$, $x + (y + z) = (x + y) + z$.
  \pause

  \textbf{Example B.} \only<6>{Is~$1 + 1 = 0$ in~$\AA$?}%
  \pause
  \only<7->{It is \hil{not the case} that~$1 + 1 = 0$ in~$\AA$. \only<8->{But also:}} \\
  \pause
  \phantom{\textbf{Example B.}} It is \hil{not the case} that~$1 + 1 \neq 0$ in~$\AA$.
  \pause

  \textbf{Example C (Anders Kock).} The generic ring is a \hil{field}:
  \[ \forall x \in \AA\_ \bigl((x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y
  \in \AA\_ xy = 1)\bigr). \]

  \pause
  \vspace*{-0.6em}
  \hil{Hence:} When verifying a coherent sequent for all rings, can without loss of
  generality assume the field condition.
\end{frame}

\begin{frame}{A selection of noncoherent sequents}
  The \hil{generic object}~$\MM$ validates:
  \begin{enumerate}
    \item $\forall x,y \in \MM\_ \neg\neg(x = y)$.
    \item $\forall x_1,\ldots,x_n \in \MM\_ \neg \forall y \in \MM\_
    y = x_1 \vee \cdots \vee y = x_n$.
  \end{enumerate}

  The \hil{generic ring}~$\AA$ validates:
  \begin{enumerate}
    \item $\forall x \in \AA\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y \in \AA\_ xy = 1)$.
    \item $\forall x \in \AA\_ \neg\neg(x = 0)$.
  \end{enumerate}

  The \hil{generic local ring}~$\AA'$ validates:
  \begin{enumerate}
    \item $\forall x \in \AA'\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y \in \AA'\_ xy = 1)$.
    \item $\neg \forall x \in \AA'\_ \neg\neg(x = 0)$.
    \item $\forall f \in \AA'[X]_{\text{degree} > 0}\_ \neg\neg \exists x \in \AA'\_ f(x) = 0$.
  \end{enumerate}
\end{frame}

\begin{frame}{An application in commutative algebra}
  \justifying
  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
  Let~$\pp$ be the \hil{generic prime ideal$^\star$} of~$A$.
  Then~$A_\pp \defeq A[\pp^{-1}]$ validates:

  {\vspace*{-1.2em}
  \setbeamercolor{block body}{bg=red!30}
  \setbeamercolor{structure}{fg=purple}
  \begin{varblock}{\textwidth}{}
    $A_\pp$ is a \hil{field}: $\forall x \in A_\pp\_ (\neg(\exists y \in A_\pp\_
    xy = 1) \Rightarrow x = 0)$.

    $A_\pp$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
    $\forall x,y \in A_\pp\_ \neg\neg(x = y) \Rightarrow x = y$.

    \mbox{$A_\pp$ is \hil{anonymously Noetherian}.}\\[-1.2em]
  \end{varblock}}

  This observation unlocks a short and conceptual proof of Grothen\-dieck's
  \hil{generic freeness lemma} in algebraic geometry.

  \mbox{\textbf{Thm.} (baby freeness) Let~$M$ be an~$A$-module. Then~\bignumber{1}
  implies~\bignumber{3}.}

  {\small
  \begin{tabular}{@{}lc@{\ }l@{}}
    \bignumber{1} $M$ is finitely generated &($\Longleftrightarrow$&
    $M_\pp$ is finitely generated) \\
    \bignumber{2} $M$ is locally free &($\Longleftrightarrow$&
    $M_\pp$ is free) \\
    \bignumber{3} $M$ is locally free \emph{on a dense open} &($\Longleftrightarrow$&
    $M_\pp$ is \emph{not not} free) \\
  \end{tabular}\par}

  \textbf{Proof.} Elementary linear algebra over~$A_\pp$. \qed
\end{frame}

\begin{frame}{A systematic source}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
    \scriptsize
    Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
    Proc.\@ of the ICM (Helsinki, 1978).
  \end{center}

  \textbf{Thm. (Nullstellensatz):} The generic~$\TT$-model~$U_T$ validates:
  For any coherent sequent~$\sigma$,
  \[
    \text{$\sigma$ holds for $U_\TT$} \quad\textcolor{purple}{\Longleftrightarrow}\quad
    \text{$\ul{\TT}/U_\TT$ proves $\sigma$.}
  \]

  \justifying
  \textbf{Thm. (universality):} The generic~$\TT$-model validates a first-order
  formula~$P$ if and only if~$P$ is intuitionistically deducible from the
  axioms of~$\TT$ and the Nullstellensatz.
\end{frame}

\begin{frame}{Arithmetic universes}
  Places where we can do mathematics (among others):
  \begin{multicols}{2}
    \begin{enumerate}
      \item $\Set$ (sets)
      \item $\Eff$ (data types) \columnbreak
      \item \mbox{$\mathrm{sSet}$ (simplicial sets)}
      \item \mbox{$\Sh(X)$ (sheaves over~$X$)}
    \end{enumerate}
  \end{multicols}
  \pause
  \vspace*{-0.6em}
  These are examples for \hil{arithmetic universes}.
  \pause

  \justifying
  \textbf{Definition.} An \emph{arithmetic universe} is a category with finite
  limits (``$\times$''), stable finite disjoint coproducts (``$\amalg$''),
  stable ef{}fective quotients (``$X/{\sim}$'') and parametrized list objects
  (``$\NN$'', ``$\mathrm{List}(X)$'').

  \textbf{Thm.} Any statement which is provable in \hil{predicative constructive
  mathematics} (no powersets, no $\varphi \vee \neg\varphi$, no
  \mbox{$\neg\neg\varphi \Rightarrow \varphi$}, no axiom of choice) is true in any
  arithmetic universe.
  \pause

  Further examples:

  \begin{enumerate}
    \addtocounter{enumi}{4}
    \item the \hil{initial} arithmetic universe
    \item the \hil{classifying} arithmetic universe for the theory of rings
  \end{enumerate}
\end{frame}
